---
title: "Simple Exploration of Heart Disease Data"
authors: "Miftaj Chowdhury, Abhi Adusumilli, Brady Seyler"
data-modified: now
format: pdf

---
Welcome, this code is written in a very beginner-friendly way, every step is explained in simple English.


# Loading Packages

```{r}
#| label: loadPackages
#| eval: true
#| echo: true
#| results: hide
# Loading useful packages
library(tidyverse)
library(ggplot2)
```


# Step 1: Reading the dataset

```{r}
#| label: readData
#| eval: true
#| echo: true
#| results: hide
# Reading the heart data into R
heart <- read_csv("heart.csv .txt", col_names = FALSE)
```


# Look at the first few lines of the dataset

```{r}
#| label: viewData
#| eval: true
#| echo: true
# Viewing the first lines of the data set
head(heart)
```


# Step 2: Adding names to each column

```{r}
#| label: columnNames
#| eval: true
#| echo: true
#| results: hide
# These names come from the original UCI Heart Disease dataset.
colnames(heart) <- c(
  "age", "sex", "cp", "trestbps", "chol", "fbs",
  "restecg", "thalach", "exang", "oldpeak",
  "slope", "ca", "thal", "target"
)
```

# Step 3: Cleaning the data

```{r}
#| label: tidyData
#| eval: true
#| echo: true
#| results: hide
#Converting the data into all numeric values
heart_clean <- heart %>%
  mutate(across(everything(), as.numeric))
#Removing unneeded rows
heart_clean <- heart_clean %>%
  filter(target <= 1)
#Fixing the names for the factor variables
heart_clean <- heart_clean %>%
  mutate(
    sex = factor(sex, labels = c("Female", "Male")),
    target = factor(target, labels = c("No Disease", "Disease")),
    cp = factor(cp),
    exang = factor(exang),
    slope = factor(slope)
  )
```

Some columns were read in as text (characters) such as "0.0".
We simply convert EVERYTHING into numbers.
heart_clean <- heart %>%
  mutate(across(everything(), as.numeric))
We are only interested in cases that have heart disease or not, so we removed all cases that were labeled with 2, 3, or 4 in the target column. 
"sex" is 0 or 1 → we rename these to "Female" and "Male"
"target" is 0 or 1 (disease indicator) in YOUR dataset,
so we rename these to "No Disease" and "Disease".

# Step 4: Simple summary statistics

```{r}
#| label: summaryStats
#| eval: true
#| echo: true
#| results: hide
#These give us a quick overview of the data.
heart_clean %>%
  summarize(
    average_age = mean(age),
    average_cholesterol = mean(chol),
    average_resting_bp = mean(trestbps)
  )
# Count how many males/females have disease/no disease
heart_clean %>% count(sex, target)
```

# Looking at tidied data

```{r}
#| label: summaryStats
#| eval: true
#| echo: true
# Viewing the tidied data
glimpse(heart_clean)
```

# Visualizations

```{r}
#| label: Plot1 
#| eval: true
#| echo: true
# Bar Graph shows information on who does and doesn't have heart disease.
ggplot(heart_clean, 
       aes(
         x = target,
         colour = sex)
       ) +
  geom_bar() +
  labs(
    title = "Heart Disease vs No Heart Disease",
    subtitle = "Sepereated by Sex",
    x = "Condition",
    y = "Number of People",
    colour = "Sex",
    alt = "Bar Chart showing how many men and women have heart disease within the data."
  ) +
  theme_minimal()
```

This plot shows the number of people in the data set that had and did not have heart disease. It additionally shows the sex proportions between who did and did not have heart disease. From the graph you can see that there are significantly more men that had heart disease than women in the data, which coincides with the table that was shown earlier. 

```{r}
#| label: Plot2 
#| eval: true
#| echo: true
# Histogram shows information on the ages of the people in the data set.
ggplot(heart_clean,
       aes(age)
       ) + 
  geom_histogram(bins = 20, fill = "blue") + 
  labs(
    title = "Distribution of Patient Ages", 
    x = "Age", 
    y = "Count",
    alt = "Histogtram showing the distribution of the ages of the people in the dataset."
    ) + 
  theme_minimal()
```

Knowing the average age of the people in the data set is important when interpreting the data, which is what this histogram shows us. The range of ages spans from about 30 to 75. The distribution of the ages is bimodal, with the two peaks being around 45 and 55. This information will help bring context to future visualizations by knowing the ages of the people in the data set. 

```{r}
#| label: Plot3 
#| eval: true
#| echo: true
# Box plot shows difference in Cholesterol levels for people with and without heart disease.
ggplot(heart_clean, 
       aes(target, chol)
       ) + 
  geom_boxplot() + 
  labs(
    title = "Cholesterol Levels by Heart Disease Status", 
    x = "Condition", 
    y = "Cholesterol",
    alt = "Box plots showing the cholesterol levels for people with and without heart disease."
    ) +
  theme_minimal()
```

This side by side box plot shows the difference in cholesterol levels for people with and without heart disease. The average cholesterol levels for people with heart disease seems to be higher than the levels for people without. While we can't confirm exact causation with just this plot, there is certainly a possibility that higher cholesterol levels could be a factor in heart disease. 

```{r}
#| label: Plot4 
#| eval: true
#| echo: true
# Box plot shows difference in resting blood pressure for people with and without heart disease.
ggplot(heart_clean, 
       aes(target, trestbps)
       ) + 
  geom_boxplot() + 
  labs(
    title = "Resting Blood Pressure by Heart Disease Status", 
    x = "Condition", 
    y = "Blood Pressure",
    alt = "Box plots showing the resting blood pressure for people with and without heart disease."
    ) +
  theme_minimal()
```

This side by side box plot shows the difference in resting blood pressure for people with and without blood pressure. The average blood pressure is almost identical for people with and without blood pressure, so this can not be used as an indicator for heart disease from this data set alone.

```{r}
#| label: Plot5 
#| eval: true
#| echo: true
# Box plot shows difference in max heart rate for people with and without heart disease. 
ggplot(heart_clean, 
       aes(target, thalach)
       ) + 
  geom_boxplot() + 
  labs(
    title = "Max Heart Rate by Heart Disease Status", 
    x = "Condition", 
    y = "Max Heart Rate",
    alt = "Box plots showing the max heart rate for people with and without heart disease."
    ) +
  theme_minimal()
```

This side by side box plot shows the difference in max heart rate for people with and without heart disease. The average max heart rate for people with heart disease seems to be less than the average heart rate for people without heart disease. This should be considered when looking into factors for heart disease.
## Relationship Between Age and Maximum Heart Rate

#| label: Plot3
#| eval: true
#| echo: true

ggplot(
heart_clean,
aes(
x = age,
y = thalach,
color = target
)
) +
geom_point(alpha = 0.6) +
labs(
title = "Age vs Maximum Heart Rate",
subtitle = "Colored by Heart Disease Status",
x = "Age",
y = "Maximum Heart Rate Achieved",
color = "Condition",
alt = "Scatter plot showing age on the x-axis and maximum heart rate on the y-axis, colored by heart disease status."
) +
theme_minimal()

This scatter plot explores the relationship between age and maximum heart rate achieved.
